Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2021-05-16 2021/emnlp/auto_cl_mdmt_v1.tex
local changes: correcting spelling, adding sacrebleu reference, adding cluster-domain proportion table

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  Building effective Neural Machine Translation models often imply accommodating diverse sets of heterogeneous data so as to optimize performance for the domain(s) of interest. Such multi-source / multi-domain adaptation problems are typically approached through instance selection or reweighting strategies, based on a static assessment of the relevance of training instances with respect to the task at hand. In this paper, we study dynamic data selection strategies that are able to automatically re-evaluate the usefulness of data samples and to evolve a data selection policy in the course of training. Based on the results of multiple experiments, we show that such methods constitute a generic framework to automatically and effectively cater to a variety of real-world situations, from supervised multi-source domain adaptation to multi-domain learning and unsupervised domain adaptation. 
\end{abstract}

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for the task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key component in training is then to mitigate the detrimental effects of a possible mismatch between these distributions. Single-source and multi-source\footnote{In this paper, multi-source DA means having multiple domains to adapt from and should not be confused with multi-source translation, where several \emph{source languages} are considered.} domain adaptation (DA) is a well-studied instance of this setting (see \citet{Chu2017comparison,Saunders21domain} for a review), and so is multi-domain (MD) learning \citep{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual machine translation \citep{Firat16multiway,Ha16towards,Johnson17google,Aharoni19massively},\fyDone{Fix ref}
% \revision{cite also \citep{Shen21source}} -> I think this is not so appropriate here
where the heterogeneity of training data not only corresponds to variations in topic, genre or register, but also in language.

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions \citep{Moore10selection,Axelrod11domain}. Finding the optimal balance of training data is however a challenging task due, for instance, to the similarity between domains/languages, but also due to regularization effects of out-of-domain data \citep{Miceli-barone17regularization}. It may also be suboptimal, as some target domains or languages might be easier to train than others. For example, improving the performance of the MT system in one domain will often hurt that of another \citep{Vanderwees17dynamic, Britz17mixing}, and improving model generalization across all domains \citep{koehn18findings} may not achieve optimily for any particular domain.

@@

In this paper, we contribute to this line of research in several ways. First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short) that simultaneously accounts for the domain adaptation and the multi-domain adaptation problems. Second, we propose a novel method of building an automated curriculum for training Multi-Domain NMT models. We evaluate our method in various single-domain and multi-domain adaptation settings. We compare our method with several contrasts including the work of \citet{Zhang19curriculum} and \citet{Wang20balancing}, which was previously only applied to multilingual MT.
% From now on, we name our method MDAC which is brief for .
Based on these experimental results, our main conclusions are that (a) using MDAC often yields overall performance that is as good as the standard fine-tuning strategy for domain adaptation; (b) MDAC can effectively handle a variety of test situations, from targeting one single domain to the full multi-domain scenario. % distributions

@@

@@

% Our method
MDAC constructs an adaptative training distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution using the REINFORCE algorithm of \citet{Williams92simple}. The evolution of $\psi$ is thus defined by:
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{1} * \displaystyle{\mathop{\sum}_{d=1}^{n_d}} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi}, \\
\end{align*}
\begingroup
\allowdisplaybreaks
where reward $R(d)$ is computed as:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t),
\end{align*}
and where we also define:
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& Update\big(\theta_{t+i-1},[x^i_j,y^i_j]_{j=1}^N\big) \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d).
\end{array}
\end{equation}
\endgroup
% \theta_{t+i-1} - \mathbf{lr_2} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta}
In these equations, $N$ denotes the size of a batch; $\mathbf{lr}_{1}$ is the learning rate of the sampling distribution; $\mathbf{lr_2}$ is the learning rate of the NMT system; $l(\theta,x,y)$ is the loss of the NMT model on sample $(x,y)$; $J^t(\theta,\lambda_t)$ is the weighted loss aggregated over $n_d$ dev-sets corresponding to the $n_d$ domains.

To compute the reward $R(d)$ of using data from domain $d$ to train the model, we simulate $k$ training steps from current checkpoint by training the model with $k$ batches sampled from $D^l(d)$ and compute the gain of the weighted dev-loss. This computation is inspired by the target prediction gain in \citep{Graves17automated}. However, where \cite{Graves17automated} used accumulated gains from the past as reward, we instead predict the usefulness of each domain for improving the future performance of the system given its current state. This is achieved by simulating a round of training with only the data from one domain. Furthermore, the choice of the parameterization of the sampling distribution in our method differs from that of \citet{Graves17automated}, although this choice is empirical.

@@

\subsection{MDAC for (multi) domain adaptation}
The setting developed in previous sections is quite general and can in principle accommodate the variety of situations mentioned above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments, we would like to better assess the true potential of MDAC is these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is MDAC a viable alternative to conventional fine-tuning? In particular, does it enable to better take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform \emph{unsupervised} (multi-)domain adaptation? \fyDone{TBContinued}
\end{itemize}
These questions are further explored in section~\ref{sec:results}. We now turn to our experimental conditions.

@@

We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.\footnote{We use truecasing and the \texttt{multibleu} script.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \cite{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \cite{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

Additional choices for the hyper-parameters of our method include the number of simulation steps $k=10$ and the learning rate $\mathbf{lr}_{data}=0.001$. We update the sampling distribution via 100 gradient descent iterations for almost all experimental settings except that for adaptation with automatic clusters (Section~\ref{ssec:clda}), where we use 20~gradient descent iterations to slow down convergence.\fyFuture{Why? computation?} We split the training into 100 short sessions that last 2000 training steps each. The choice of those hyper-parameters is mostly heuristic except for the learning rate $\mathbf{lr}_{data}$ which is optimized via grid search over a range of values $[0.001,0.0025,0.005]$.

@@

In a fourth experiment (\emph{few shot adaptation}), given training data in 6 domains and a small development set of a new domain, (\domain{News} in our case), we aim to build an NMT model which performs well for the unseen domain.

Finally, in the \emph{unsupervised domain adaptation task}, we cluster all available training data into 30 clusters using the KNN algorithm in the same way as in \citep{Tars18multidomain}, then adapt these clusters to one of 6 domains using the corresponding in-domain dev set. We compare MDAC to DDS for each our 6 test sets.

@@

We now compare the best MDAC system (using $\vlambda_{t=0} = \vlambda_0$) to full fine-tuning. According to Table~\ref{tab:da}, fine-tuning is better for large domains such as \domain{med} and \domain{law}, while MDAC outperforms fine-tuning by approximately 0.5 BLEU for \domain{bank} and 1.0 BLEU for \domain{rel}. This indicates that for small domains, out-of-domain data helps improve the generalization and that MDAC is able to exploit this advantage by balancing the in-domain and the out-of-domain training data instead of edging out the out-of-domain training data as in fine-tuning. The results for DDS display similar trends, but are always outperformed by MDAC. The behaviour of CL, which does only well the large domain \domain{med} lag somewhat behind.

@@

A second observation is that DDS is again outperformed by MDAC by a significant margin of 1.5 BLEU on average; the only domain where DDS does (much) better is \domain{med}. Figure~\ref{fig:sampling}, where we plot the evolution of the training mixture in the course of training sessions, helps understand the difference between the two methods. For DDS (Figure~\ref{fig:DDS}), the proportion of 6 domains quicly converges towards a bi-domain mode in which only \domain{med} and \domain{rel} have significant probability -- hence the good performance on the former domain. In contrast, the proportion of 6~domains computed by MDAC evolves more smoothly; small domains such as \domain{bank}, \domain{it}, \domain{talk} and \domain{rel} receive a larger proportion of training data in the early stages; their weights then slowly decrease as larger domains such as \domain{med} and \domain{law} increase their share. This however only happends at the end of training, when the NMT models might have been already close to their optimal performance for the small domains.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*{10}{r|}} \cline{1-2} \cline{4-11}
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{news}}& \hfill &domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean}  \\ 
\cline{1-2} \cline{4-11}
    \multicolumn{2}{|c|}{\sl Unseen domain} & &\multicolumn{8}{c|}{\sl Training with 30 clusters} \\ 
\cline{1-2} \cline{4-11}
    \system{Mixed-0}      & 25.7 & &\system{DDS($\vlambda^*, \vlambda_d$)}&38.3&60.1&50.3&35.8&49.1&90.1&53.9\\
    \system{Mixed-0.25} & 25.8 & &\system{MDAC($\vlambda^*, \vlambda_d$)}&39.2$^*$&61.6$^*$&52.0$^*$&38.2$^*$&49.1&89.7&55.0$^*$\\ \cline{4-11}
    \system{Mixed-0.5}   &26.5\\
    \system{Mixed-0.75} &26.8\\
    \system{Mixed-1} &26.9 \\
    \cline{1-2}
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \system{MDAC($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \cline{1-2}
  \end{tabular}
  \caption{Few shot adaptation (left) and unsupervised adaptation (right). For a given line, each column correspond to one distinct system. ($^*$) MDAC is significantly better than DDS.}
  \label{tab:unsupervised-da}
\end{table*}

\subsection{Unseen domain}\label{ssec:uda}
The left part of Table~\ref{tab:unsupervised-da} displays the performance on the unseen domain \domain{news} for NMT systems trained with the mixtures $\lambda^l \in \big[ \lambda_0, \lambda_{0.25}, \lambda_{0.5}, \lambda_{0.75}, \lambda_{1.0}\big]$ and wiht dynamic data selection (MDAC and DDS). These systems have insignificant differences in BLEU, showing that dynamic mixture does not improve the robustness of the NMT system against unseen domains. However, the fact that the performance of MDAC and DDS is close to the best performance is a sign that they can also apply in such challenging situation. % the robustness of the NMT model.

\subsection{Automatic clustering}\label{ssec:clda}
The right part of Table~\ref{tab:unsupervised-da} reports the performance of NMT systems adapted to each domain. In comparison to Section~\ref{ssec:da}, the training data is distributed in 30 automatic clusters instead of the 6 original domains. Spliting the train data in small groups provides the learner with extra degrees of freedom when selecting the best distribution. However, as these clusters are built automatically, they are more noisy in nature. According to the results in Table~\ref{tab:unsupervised-da}, this scenario is hard both for DDS and MDAC which perform much worse than for the supervised DA setting. This again signals the importance of initialization: analyzing the clustering, we find that the data for \domain{rel} mostly correspond to one single cluster. Using a uniform initialization, this cluster starts with a very small weight, and never succeeds to yield the (good) performance observed in the DA settting.

@@

Domain adaptation is an old problem that has been studied from many angles both for SMT and NMT. A survey of supervised and unsupervised DA for NMT is in \citet{Chu18asurvey}, where the authors distinguish between data-centric and model-centric DA, a view also adopted in the more recent review of \citet{Saunders21domain}. Our approach to DA in this paper falls under the former category. We refer interested readers to these references and do not discuss DA any further.

Multi-domain adaptation (MDA) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDA combine one or several ingredients: (a) the specialization of data representations \citep{Kobus17domaincontrol} or of sub-networks \citep{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \citep{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \citet{Jiang19multidomain}. Unsupervised MDA is studied in \citet{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive / dynamic data selection take inspiration from \citet{Bengio09curriculum}, where the notion of curriculum learning is introduced. CL relies on the notion of ``easiness'' of a sample to schedule the presentation of training data so that the easiest examples are presented first and the hardest, last. Various ways to automate CL using the framework of multi-armed bandits are explored in \citep{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for domain adaptive / multi-domain / multi-lingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}
For instance, \citet{Zhang19curriculum} study supervised DA and propose a curriculum approach which progressively augments the training data: in the early stages, only in-data is used, while shards containing less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is somehow opposite to the recommendations of \citet{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively focuses on the in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\citet{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to multilingual NMT in \citep{Kumar21learning}. \citet{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative criterium for stopping the training. More related to our problems, \citet{Wang20learning-multi} adapt CL for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the development set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}

\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform a variety of standard adaptation tasks for machine translation, ranging from the conventional supervised domain adaptation to multi-domain adaptation and few shot DA. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during  the course of training, could be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA that existing heuristics, and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approachs to dynamic data selection.

In the future, we intend to continue develop this approach and improve its effectiveness. One issue that we have left unaddressed is reward normalization, which is especially important in the early stages of training \cite{Kumar19reinforcement}. Another area where we need to progress is the unsupervised learning setting of Section~\ref{ssec:clda}, where our results still lag way behind that of supervised DA - this might be due to the inability of our simplistic optimization strategy to handle situations where the number of clusters is too large.
% slow down learning during warm up or normalize reward; handle large number of classes;
% too much importance to the dev ? randomize ?

>>>>

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  Building effective Neural Machine Translation models often implies accommodating diverse sets of heterogeneous data so as to optimize performance for the domain(s) of interest. Such multi-source / multi-domain adaptation problems are typically approached through instance selection or reweighting strategies, based on a static assessment of the relevance of training instances with respect to the task at hand. In this paper, we study dynamic data selection strategies that are able to automatically re-evaluate the usefulness of data samples and to evolve a data selection policy in the course of training. Based on the results of multiple experiments, we show that such methods constitute a generic framework to automatically and effectively cater to a variety of real-world situations, from supervised multi-source domain adaptation to multi-domain learning and unsupervised domain adaptation. 
\end{abstract}

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for the task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key component in training is then to mitigate the detrimental effects of a possible mismatch between these distributions. Single-source and multi-source\footnote{In this paper, multi-source DA means having multiple domains to adapt from and should not be confused with multi-source translation, where several \emph{source languages} are considered.} domain adaptation (DA) is a well-studied instance of this setting (see \citet{Chu2017comparison,Saunders21domain} for a review), and so is multi-domain (MD) learning \citep{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual machine translation \citep{Firat16multiway,Ha16towards,Johnson17google,Aharoni19massively},\fyDone{Fix ref}
% \revision{cite also \citep{Shen21source}} -> I think this is not so appropriate here
where the heterogeneity of training data not only corresponds to variations in the topic, genre, or register but also in language.

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions \citep{Moore10selection,Axelrod11domain}. Finding the optimal balance of training data is however a challenging task due, for instance, to the similarity between domains/languages, but also due to regularization effects of out-of-domain data \citep{Miceli-barone17regularization}. It may also be suboptimal, as some target domains or languages might be easier to train than others. For example, improving the performance of the MT system in one domain will often hurt that of another \citep{Vanderwees17dynamic, Britz17mixing}, and improving model generalization across all domains \citep{koehn18findings} may not achieve optimally for any particular domain.

@@

In this paper, we contribute to this line of research in several ways. First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short) that simultaneously accounts for the domain adaptation and the multi-domain adaptation problems. Second, we propose a novel method of building an automated curriculum for training Multi-Domain NMT models. We evaluate our method in various single-domain and multi-domain adaptation settings. We compare our method with several contrasts, including the work of \citet{Zhang19curriculum} and \citet{Wang20balancing}, which was previously only applied to multilingual MT.
% From now on, we name our method MDAC, which is brief for .
Based on these experimental results, our main conclusions are that (a) using MDAC often yields overall performance that is as good as the standard fine-tuning strategy for domain adaptation; (b) MDAC can effectively handle a variety of test situations, from targeting one single domain to the full multi-domain scenario. % distributions

@@

% === The part on multidomain training has been moved to the end.

@@

% Our method
MDAC constructs an adaptative training distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution using the REINFORCE algorithm of \citet{Williams92simple}. The evolution of $\psi$ is thus defined by:
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{1} * \displaystyle{\mathop{\sum}_{d=1}^{n_d}} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi}, \\
\end{align*}
\begingroup
\allowdisplaybreaks
where reward $R(d)$ is computed as:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t),
\end{align*}
and where we also define:
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& Update\big(\theta_{t+i-1},[x^i_j,y^i_j]_{j=1}^N\big) \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d).
\end{array}
\end{equation}
\endgroup
% \theta_{t+i-1} - \mathbf{lr_2} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta}
In these equations, $N$ denotes the size of a batch; $\mathbf{lr}_{1}$ is the learning rate of the sampling distribution; $l(\theta,x,y)$ is the loss of the NMT model on sample $(x,y)$; $J^t(\theta,\lambda_t)$ is the weighted loss aggregated over $n_d$ dev-sets corresponding to the $n_d$ domains.

To compute the reward $R(d)$ of using data from domain $d$ to train the model, we simulate $k$ training steps from the current checkpoint by training the model with $k$ batches sampled from $D^l(d)$ and compute the gain of the weighted dev-loss. This computation is inspired by the target prediction gain in \citep{Graves17automated}. However, where \cite{Graves17automated} used accumulated gains from the past as rewards, we instead predict the usefulness of each domain for improving the system's future performance given its current state. This is achieved by simulating a round of training with only the data from one domain. Furthermore, the choice of the parameterization of the sampling distribution in our method differs from that of \citet{Graves17automated}, although this choice is empirical.

@@

\subsection{MDAC for (multi) domain adaptation}
The setting developed in previous sections is quite general. It can, in principle, accommodate the variety of situations mentioned above and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments, we would like to assess better the true potential of MDAC in these settings and seek to answer the following questions experimentally:
\begin{itemize}
\item is MDAC a viable alternative to conventional fine-tuning? In particular, does it enable better to take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform \emph{unsupervised} (multi-)domain adaptation? \fyDone{TBContinued}
\end{itemize}
These questions are further explored in section~\ref{sec:results}. We now turn to our experimental conditions.

@@

We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \cite{Papineni02bleu}.\footnote{We use truecasing and \texttt{sacrebleu}\citep{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \cite{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \cite{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

Additional choices for the hyper-parameters of our method include the number of simulation steps $k=10$ and the learning rate $\mathbf{lr}_{data}=0.001$. We update the sampling distribution via 100 gradient descent iterations for almost all experimental settings except that for adaptation with automatic clusters (Section~\ref{ssec:clda}), where we use 20~gradient descent iterations to avoid converging to degenerate distributions.\fyFuture{Why? computation?} We split the training into 100 short sessions that last 2000 training steps each. The choice of those hyper-parameters is mostly heuristic except for the learning rate $\mathbf{lr}_{data}$ which is optimized via grid search over a range of values $[0.001,0.0025,0.005]$.

@@

In a fourth experiment (\emph{few-shot adaptation}), given training data in 6 domains and a small development set of a new domain, (\domain{News} in our case), we aim to build an NMT model which performs well for the unseen domain.

Finally, in the \emph{unsupervised domain adaptation task}, we cluster all available training data into 30 clusters using the KNN algorithm in the same way as in \citep{Tars18multidomain}, then adapt these clusters to one of 6 domains using the corresponding in-domain dev set. We compare MDAC to DDS for each of our 6 test sets.

@@

We now compare the best MDAC system (using $\vlambda_{t=0} = \vlambda_0$) to full fine-tuning. According to Table~\ref{tab:da}, fine-tuning is better for large domains such as \domain{med} and \domain{law}, while MDAC outperforms fine-tuning by approximately 1.2 BLEU for \domain{bank} and 1.0 BLEU for \domain{rel}. This indicates that for small domains, out-of-domain data helps improve the generalization and that MDAC is able to exploit this advantage by balancing the in-domain and the out-of-domain training data instead of edging out the out-of-domain training data as in fine-tuning. The results for DDS display similar trends but are always outperformed by MDAC. The behavior of CL, which does only well on the large domain \domain{med}, lag somewhat behind.

@@

A second observation is that DDS is again outperformed by MDAC by a significant margin of 1.5 BLEU on average; the only domain where DDS does (much) better is \domain{med}. Figure~\ref{fig:sampling}, where we plot the evolution of the training mixture in the course of training sessions, helps understand the difference between the two methods. For DDS (Figure~\ref{fig:DDS}), the proportion of 6 domains quickly converges towards a bi-domain mode in which only \domain{med} and \domain{rel} have significant probability -- hence the good performance on the former domain. In contrast, the proportion of 6~domains computed by MDAC evolves more smoothly; small domains such as \domain{bank}, \domain{it}, \domain{talk} and \domain{rel} receive a larger proportion of training data in the early stages; their weights then slowly decrease as larger domains such as \domain{med} and \domain{law} increase their share. This, however, only happens at the end of the training, when the NMT models might have already reached their optimal performance for the small domains.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*{10}{r|}} \cline{1-2} \cline{4-11}
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{news}}& \hfill &domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean}  \\ 
\cline{1-2} \cline{4-11}
    \multicolumn{2}{|c|}{\sl Unseen domain} & &\multicolumn{8}{c|}{\sl Training with 30 clusters} \\ 
\cline{1-2} \cline{4-11}
    \system{Mixed-0}      & 25.7 & &\system{DDS($\vlambda^*, \vlambda_d$)}&38.3&60.1&50.3&35.8&49.1&90.1&53.9\\
    \system{Mixed-0.25} & 25.8 & &\system{MDAC($\vlambda^*, \vlambda_d$)}&39.2$^*$&61.6$^*$&52.0$^*$&38.2$^*$&49.1&89.7&55.0$^*$\\ \cline{4-11}
    \system{Mixed-0.5}   &26.5\\
    \system{Mixed-0.75} &26.8\\
    \system{Mixed-1} &26.9 \\
    \cline{1-2}
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \system{MDAC($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \cline{1-2}
  \end{tabular}
  \caption{Few shot adaptation (left) and unsupervised adaptation (right). For a given line, each column correspond to one distinct system. ($^*$)MDAC is significantly better than DDS.}
  \label{tab:unsupervised-da}
\end{table*}

\subsection{Unseen domain}\label{ssec:uda}
The left part of Table~\ref{tab:unsupervised-da} displays the performance on the unseen domain \domain{news} for NMT systems trained with the mixtures $\lambda^l \in \big[ \lambda_0, \lambda_{0.25}, \lambda_{0.5}, \lambda_{0.75}, \lambda_{1.0}\big]$ and with dynamic data selection (MDAC and DDS). These systems have insignificant differences in BLEU, showing that dynamic mixture does not improve the robustness of the NMT system against unseen domains. However, the fact that the performance of MDAC and DDS is close to the best performance is a sign that they can also apply in such challenging situations. % the robustness of the NMT model.

\subsection{Automatic clustering}\label{ssec:clda}
The right part of Table~\ref{tab:unsupervised-da} reports the performance of NMT systems adapted to each domain. In comparison to section~\ref{ssec:da}, the training data is distributed in 30~clusters instead of the 6 original domains. Splitting the train data into small groups provides the learner with extra degrees of freedom when selecting the best distribution. However, as these clusters are built automatically, they are noisier in nature. According to the results in Table~\ref{tab:unsupervised-da}, this scenario is hard both for DDS and MDAC, which perform much worse than for the supervised DA setting. This again signals the importance of initialization: analyzing the clustering, we find that the data for \domain{rel} mostly correspond to one single cluster. Using a uniform initialization, this cluster starts with a very small weight and never succeeds in yielding the (good) performance observed in the DA setting.

@@

Domain adaptation is an ancient and vexing problem that has been studied from many angles, both for SMT and NMT. A survey of supervised and unsupervised DA for NMT is in \citet{Chu18asurvey}, where they distinguish between data-centric and model-centric DA, a view also adopted in the more recent review of \citet{Saunders21domain}. Our approach to DA in this paper clearly falls under the former category. We refer interested readers to these references and do not discuss DA any further.

Multi-domain adaptation (MDA) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDA typically combine one or several ingredients: (a) the specialization of data representations \citep{Kobus17domaincontrol} or of sub-networks \citep{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \citep{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification, e.g., \ \citet{Jiang19multidomain}. Unsupervised MDA is studied in \citet{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive/dynamic data selection take inspiration from \citet{Bengio09curriculum}, where the notion of curriculum learning is initially introduced. CL relies on the notion of the ``easiness'' of a sample to schedule the presentation of the training data so that the easiest examples presented first and the hardest, last. Various ways to automate CL using the framework of multi-armed bandits are notably explored in \citep{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for domain adaptive / multi-domain / multilingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}

For instance, \citet{Zhang19curriculum} study supervised domain adaptation and propose a curriculum approach which progressively extends the data used in training: in the early stages, only in-data is used, while shards containing less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is somehow opposite to the recommendations of \citet{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively focuses on the in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\citet{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this study, complexity corresponds to difficulty levels and which are binned based on contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to the training of a multilingual NMT system \citep{Kumar21learning}. \citet{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative criterium for stopping the training. More related to our problems, \citet{Wang20learning-multi} adapt curriculum learning for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization (derivative-free) techniques. Each step consists of (a) weighting the instances based on relevance features, (b) fine-tuning a pretrained model using the (weighted) training set, and is applied iteratively to generate a sequence of models. The one that maximizes the development set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
This study presents a generic framework to perform various standard adaptation tasks for machine translation, ranging from the conventional supervised domain adaptation to multi-domain adaptation and few-shot DA. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during training, could be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.

In the future, we intend to improve this approach by solving several issues. One issue that we have left unaddressed is reward normalization, which is especially important in the early stages of training \cite{Kumar19reinforcement}. We also would like to improve the unsupervised learning setting of Section~\ref{ssec:clda}, where our results still lag behind that of supervised DA - this might be due to the inability of our simplistic optimization strategy to handle situations where the number of clusters is too large.

% slow down learning during warm up or normalize reward; handle large number of classes;
% too much importance to the dev ? randomize ?
